Proof equivalence in \capsabbrev{mll} is \capsabbrev{pspace}-complete

### 2015-02-02 main/LMCS.tex
Introduction

\section{Introduction}
\noindent
Sequent calculus was originally introduced by Gentzen as a way to better understand
the properties of natural deduction -- in particular the eliminability of cut. In this
view, a sequent calculus proof gives a recipe for the construction of a natural deduction
proof.

Linear logic was formulated as a sequent calculus system, and doesn’t have a corresponding
natural deduction in the classical sense. The role of natural deduction is taken instead by
proof nets. Each sequent calculus proof gives rise to a proof net. Furthermore, at least in
the case of multiplicative linear logic without units, proof nets are canonical in the sense
that two sequent calculus proofs give rise to the same proof net if and only if they are
equivalent.

Proof nets do not work so well when the logical units 1 and $\bot$ are included. Although
proof nets can still be defined in this case, they are no longer canonical. It has not

\cite{Lambek-1968}

\cite{Girard-1987,Danos-Regnier-1989}
\cite{Trimble-1994,Blute-Cockett-Seely-Trimble-1996}
\cite{Koh-Ong-1999}
\cite{Lamarche-Strassburger-2006}
\cite{Hughes-2012-categories,Hughes-2012-nets}

\cite{Hughes-vanGlabbeek-2005}
\cite{Heijltjes-2011}

\cite{Demaine-Hearn-2008}

\cite{ReconfigurationProblems}
\cite{Kanovich-1992,Lincoln-Winkler-1994}.

@@

>>>>

\section{Introduction}
\noindent
Sequent calculus was originally introduced by Gentzen as a way to better understand
the properties of natural deduction -- in particular the eliminability of cut. In this
view, a sequent calculus proof gives a recipe for the construction of a natural deduction
proof.
%
Linear logic was formulated as a sequent calculus system, and doesn't have a corresponding
natural deduction in the classical sense. The role of natural deduction is taken instead by
proof nets. Each sequent calculus proof gives rise to a proof net. Furthermore, at least in
the case of multiplicative linear logic without units, proof nets are canonical in the sense
that two sequent calculus proofs give rise to the same proof net if and only if they are
equivalent.

Proof nets do not work so well when the logical units 1 and $\bot$ are included. Although
proof nets can still be defined in this case, they are no longer canonical. Firstly the
translation from proofs to proof nets is not canonical, so a proof may be translated
to several possible proof nets with no privileged choice. Also, equivalent proofs are not
in general translated to the same possible proof nets. So proof nets lose their principal
advantage, compared to proofs, of being canonical, and are subject to a non-trivial
equivalence relation.

Equivalence of proofs (in the sense we are interested in here) was first considered
by Lambek \cite{Lambek-1968}, who introduced the idea of category-theoretic logic and used
it to define a natural notion of equivalence of proofs.

Proof nets were introduced at the birth of linear logic by Girard \cite{Girard-1987}, and
the correctness criterion was simplified by Danos and Regnier \cite{Danos-Regnier-1989}.
These proof nets did not include the logical units, and are canonical for multiplicative linear logic.
%
Canonical proof nets also exist for larger fragments of linear logic, such as the combined multiplicative-additive fragment without units \cite{Hughes-vanGlabbeek-2005}, and the additive fragment, including the additive units \cite{Heijltjes-2011}. Canonical proof nets may also be defined for \emph{intuitionistic} multiplicative linear logic with units.

Proof nets incorporating units were defined by Blute, Cockett, Seely and Trimble \cite{Trimble-1994,Blute-Cockett-Seely-Trimble-1996}, though these proof nets cease to be canonical. The `joker games' of Koh and Ong \cite{Koh-Ong-1999} have a very similar structure to proof nets, and include units. Another explicit treatment of proof nets for \capsabbrev{mll} with units is given by Lamarche and Stra\ss burger \cite{Lamarche-Strassburger-2006}, and later a particularly simple and elegant version by Hughes \cite{Hughes-2012-categories,Hughes-2012-nets}.

The idea of the present work is to look at proof equivalence through the lens of computational
complexity, and use this to settle the question of whether there can be canonical proof nets
for \capsabbrev{mll} with units. The canonical proof nets for \capsabbrev{mll} without units
give an efficient decision procedure for proof equivalence: to decide whether two such proofs
are equivalent, it is sufficient to translate both to proof nets and compare the proof nets
for equality.

The corresponding decision procedure for \capsabbrev{mll} \emph{with} units is shown to be
\capsabbrev{pspace}-complete, which is generally supposed to preclude the existence of a
polynomial-time algorithm for this problem. So there can be no canonical proof nets in the
usual sense: if the proof nets are syntactically equal just when their corresponding proofs
are equivalent, then the translation from proofs to proof nets must be intractable.

@@

Our proof of \capsabbrev{pspace}-completeness uses the \emph{nondeterministic constraint logic} of \cite{Hearn-Demaine-2005,Demaine-Hearn-2008,GamesPuzzlesAndComputation}. This is a graph rewriting
system introduced as a tool for use in complexity proofs, originally for games and puzzles.

It is not uncommon for an \capsabbrev{np}-complete problem to have an associated reconfiguration problem that is \capsabbrev{pspace}-complete \cite{ReconfigurationProblems}; an example of this is \capsabbrev{sat}-reconfiguration. \capsabbrev{mll} proof equivalence may be regarded as the reconfiguration problem associated with \capsabbrev{mll} proof search, which is \capsabbrev{np}-complete \cite{Kanovich-1992,Lincoln-Winkler-1994}.
